---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
Neil Caren
 
###Loading and Processing Data

Load necessary library files and data
```{r}
library(lattice)
activityData <- read.csv("activity.csv", stringsAsFactors = FALSE)
```

Summary of raw data

```{r}
summary(activityData)
```
####What is the mean total number of steps taken per day ?
Calculate the total number of steps taken eacch day by utiliaing lapply and split functions
break list returned by lapply into a vector to plot

```{r}
totalStepsByDate <- lapply(split(activityData$steps, activityData$date),sum)
totalStepsByDateVector <- unlist(totalStepsByDate)
```

Plot a histogram of Steps to get a sense of frequency and distribution.  Also added a rug for further clarity

```{r}
hist(totalStepsByDateVector)
rug(totalStepsByDateVector)
```

Calculate and report the mean and median of the total number of steps taken per day

```{r}
stepsMean <- round(mean(totalStepsByDateVector, na.rm=TRUE),2)
stepsMedian <- round(median(totalStepsByDateVector, na.rm = TRUE),2)
```

The mean number of steps per day = `r stepsMean` and the mediam = `r stepsMedian`


###What is the average daily activity pattern?

calculate the average number of steps taken each interval by utiliaing lapply and split functions
break list returned by lapply into a vector to plot a time series

```{r}
stepsByInterval <- lapply(split(activityData$steps, activityData$interval),mean, na.rm=TRUE)
stepsByIntervalVector <- unlist(stepsByInterval)
plot(names(stepsByIntervalVector), stepsByIntervalVector, type="l")
```

determine which 5 min interval contains the highest number of steps
```{r checkMax, echo=FALSE}
maxSteps <- round(max(stepsByIntervalVector),2)
maxInterval <- which.max(stepsByIntervalVector)
```

###The interval with the highest average steps is `r names(maxInterval)`


###Inputting missing values


```{r}
#copy activityData to cleanDataSet
cleanDataSet = activityData
#create a vector of NAs from activityData
vectorOfNAs <- activityData[(is.na(activityData$steps)),]

#loop through vectorOfNAs and use row name to update cleanDataSwt with average steps for that interval 
for (x in  1:nrow(vectorOfNAs)) {
        index = row.names(vectorOfNAs[x,])
        y <- as.character(vectorOfNAs[x,3])
        cleanDataSet[index,1] <- stepsByIntervalVector[y]
}
totNAs <- nrow(vectorOfNAs)
```


###The total number of NAs is `r totNAs`


Show summaries of before and after datasets to see change

```{r}
summary(activityData)
summary(cleanDataSet)
```


calculate the total number of steps taken eacch day by utiliaing lapply and split functions
break list returned by lapply into a vector to plot

```{r}

totalStepsByDate <- lapply(split(cleanDataSet$steps, cleanDataSet$date),sum)
totalStepsByDateVector <- unlist(totalStepsByDate)
```

Plot a histogram of Steps to get a sense of frequency.  Also added a rug for further clarity

```{r}
hist(totalStepsByDateVector)
rug(totalStepsByDateVector)
```


calculate the average number of steps taken each interval by utiliaing lapply and split functions
break list returned by lapply into a vector to plot a time series

```{r}
stepsByInterval <- lapply(split(cleanDataSet$steps, cleanDataSet$interval),mean, na.rm=TRUE)
stepsByIntervalVector <- unlist(stepsByInterval)
plot(names(stepsByIntervalVector), stepsByIntervalVector, type="l")
```

```{r checkMax2, echo=FALSE}
maxSteps <- round(max(stepsByIntervalVector),2)
maxInterval <- which.max(stepsByIntervalVector)
```

```{r}
stepsMean <- round(mean(totalStepsByDateVector),2)
stepsMedian <- round(median(totalStepsByDateVector),2)
```

The mean number of steps per day = `r stepsMean` and the mediam = `r stepsMedian`

###The interval with the highest average steps (`r maxSteps`) is `r names(maxInterval)`

###I  don't see any difference when filling in NAs mostly due to using exising averages as replacement values



###Are there differences in activity patterns between weekdays and weekends?

```{r}
#create POSIXt variable 
cleanDataSet$dtfmt <- strptime(cleanDataSet$date,"%Y-%m-%d")
#create day-of-week factor variable
cleanDataSet$dow <- factor(c("weekend"), levels = c("weekday", "weekend"))

#populate day of week 
for(x in 1:nrow(cleanDataSet)){
        if (weekdays(cleanDataSet[x,4], abbreviate = TRUE) == "Sat" || weekdays(cleanDataSet[x,4],abbreviate = TRUE) == "Sun")
                cleanDataSet[x,5] <- "weekend"
         else
                cleanDataSet[x,5] <- "weekday"
}

#plot timeseries weekend vs. weekday
xyplot(cleanDataSet$steps ~ cleanDataSet$interval | cleanDataSet$dow, layout = c( 1, 2), type = c("l", "l"))
```

